---
output: github_document
bibliography: ["readme.bib"]
---

```{r rmd-setup, include = FALSE}
library(prefio)
knitr::opts_chunk$set(fig.path = "man/figures/")
```

# prefio

Package website: https://fleverest.github.io/prefio/.

<!-- badges: start -->
[![R-CMD-check](https://github.com/fleverest/prefio/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fleverest/prefio/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Overview

Ordinal Preference datasets are used by many research communities including, but not limited to, those who work with recommender systems, computational social choice, voting systems and combinatorial optimization.

The **prefio** R package provides a set of functions which enable users to perform a wide range of preference analysis tasks, including preference aggregation, pairwise comparison summaries and convenient IO operations. This makes it easier for researchers and other professionals to perform common data analysis and preprocessing tasks with such datasets.


## Installation

<!--
The package may be installed from CRAN via

```{r, eval = FALSE}
install.packages("prefio")
```
-->

The development version can be installed via
```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("fleverest/prefio")
```


## Usage

**prefio** provides a convenient interface for processing data from tabular
formats as well as sourcing data from one of the unified
[PrefLib formats](https://www.preflib.org/format/), including a convenient
method for downloading data files directly from PrefLib to your R session.

#### Processing tabular data

Preference data can come in many forms. Commonly preference data will be
either represented in either *long*-format with each row corresponding to a
particular *ranking* chosen for a single *item*:, e.g:

```{r echo = FALSE, results = 'asis'}
long <- data.frame(
  ID = rep(1:3, each = 3),
  ItemName = LETTERS[rep(1:3, 3)],
  Rank = c(1, 2, 3, 3, 2, 1, 2, 1, 3)
)
knitr::kable(long,
             caption = "Three orderings on items {A, B, C} in long-format.")
```

This data can be converted from a `data.frame` into a `preferences` object:

```{r}
long <- data.frame(
  ID = rep(1:3, each = 3),
  ItemName = LETTERS[rep(1:3, 3)],
  Rank = c(1, 2, 3, 3, 2, 1, 2, 1, 3)
)
prefs <- preferences(long,
                     format = "long",
                     id = "ID",
                     item = "ItemName",
                     rank = "Rank")
print(prefs)
```

Another way of tabulating orderings is with each unique ordering on a single
row, with each column representing the rank given to a particular item:

```{r echo = FALSE, results = 'asis'}
rankings <- matrix(c(
  1, 2, 3,
  3, 2, 1,
  2, 1, 3
  ),
  nrow = 3,
  byrow = TRUE
)
colnames(rankings) <- LETTERS[1:3]
knitr::kable(rankings,
             caption = paste0("Three orderings on items ",
                              "{A, B, C} in a \"rankings\" format."))
```

This data can be converted from a `data.frame` into a `preferences` object:

```{r}
rankings <- matrix(c(
  1, 2, 3,
  3, 2, 1,
  2, 1, 3
  ),
  nrow = 3,
  byrow = TRUE
)
colnames(rankings) <- LETTERS[1:3]
prefs <- preferences(rankings,
                     format = "ranking")
print(prefs)
```

#### Reading from PrefLib

The [Netflix Prize](https://en.wikipedia.org/wiki/Netflix_Prize) was a
competition devised by Netflix to improve the accuracy of its recommendation
system. To facilitate this they released ratings about movies from the users of
the system that have been transformed to preference data and are available from
[PrefLib](https://www.preflib.org/data/ED/00004/), [@Bennett2007]. Each data set
comprises rankings of a set of 3 or 4 movies selected at random. Here we
consider rankings for just one set of movies to illustrate the functionality of
**prefio**.

PrefLib datafiles such as these can be downloaded on-the-fly by specifying the
argument `from_preflib = TRUE` in the `read_preflib` function:

```{r}
netflix <- read_preflib("netflix/00004-00000138.soc", from_preflib = TRUE)
head(netflix)
```

Each row corresponds to a unique ordering of the four movies in the dataset.
The number of Netflix users that assigned each ordering is given in the
`frequencies` column. In this case, the most common ordering (with 68 voters
specifying the same preferences) is the following:

```{r}
print(netflix$preferences[1], width = 100)
```


#### Writing to Preflib formats

**prefio** provides a convenient interface for writing preferential datasets to
PrefLib formats. To aid the user, the `preferences()` function automatically
calculates metrics of the dataset which are required for producing valid PrefLib
files. For example, we can write our `prefs` from earlier:

```{r}
write_preflib(prefs)
```

Note that this produces four warnings. Each warning corresponds to a field which
is required by the official PrefLib format, but may not be necessary for
internal use-cases. If your goal is to publish some data to PrefLib, these
warnings must be resolved.


## References
